% !TEX root = prelim-paper.tex

\section{Tile-based Editor Calculus}\label{sec:formalism}

We now precisely characterize tile-based editing
as a calculus called \ty.
The punchline of \ty~ (Section \ref{sec:actions})
is the action judgment
$\performAction{\editState_1}{\action}{\editState_2}$,
which sends edit state $\editState_1$ to $\editState_2$
via action $\action$; and its governing invariants,
which state that every action-reachable edit state
can be parsed into a well-formed program term.

Edit states $Z$ follow a variant of the well-known \emph{zipper}
pattern \cite{zipper}, which decomposes a structure of interest
into a focused substructure $B$ and its surrounding context $M$.
In this work, we refer to $B$ as the \emph{subject}
and $M$ as the \emph{frame}.
\[
\arraycolsep=4pt
\begin{array}{rlrl}
  \text{zipper} & \zip & ::= & \zipper{\subject}{\zframe}
\end{array}
\]
Our design diverges from prior art in our pursuit
of linear text-like interactivity.
Prior zipper designs are strictly hierarchical and ask
the client to navigate in a structure-aware two-dimensional
fashion: left and right to traverse siblings,
up and down to traverse parent-child relations.
In contrast, we aim to support a one-dimensional interaction
model:
it should be possible to navigate freely by moving solely
left and right \note{might need to clarify or change wording
here to address comparison with ordered traversals of trees};
furthermore it should be possible to specify
arbitrary subranges of a one-dimensional serialization
of the total structure.
Toward these ends, actions in \ty~ rely
on a form of incremental unparsing and reparsing by which
hierarchical structures may be ``flattened'' as needed
to support linear traversal and partial selection.

Our presentation in this section builds upon the term and tile syntax
described earlier in Figure \ref{fig:term-tile-syntax} of
Section \ref{sec:overview}.
\begin{enumerate}
  \item[\textbf{\ref{sec:shards-and-pieces}}]
  We will begin by formally introducing shards
  and describing their relation to tiles via disassembly
  and reassembly.

  \item[\textbf{\ref{sec:tips-and-terms}}]
  We will then describe how tiles and shards relate to
  terms. In particular, we will show that
  we can successfully parse a sequence of tiles
  into a term provided that they ``fit'' together in
  certain ways.
  \note{slightly misleading in that the ``fit'' property
  must hold for all disassemblages of the tiles}

  \item[\textbf{\ref{sec:subjects-and-frames}}]
  As a result, we need not limit ourselves to
  the highly structured terms
  but can instead organize our zippers around the
  more flexible tiles and shards.

  \item[\textbf{\ref{sec:actions}}]
  This paves the way for linear navigation and
  flexible selection when we define actions operating
  on zippers.
  Meanwhile, by taking care to
  maintain proper ``fit'' and shard assemblage along the
  way, actions ensure that reachable edit states can
  be successfully parsed into a program term.

\end{enumerate}


% We will begin in Section \ref{sec:terms-tiles-tips} by
% describing how tiles can be disassembled into shards,
% and how a sequence of tiles and shards is parseable
% into a well-formed term provided they ``fit'' together;
% in other words, \ty~ edit states and actions need not
% deal directly with the term syntax to enforce well-formedness.
% Having introduced the ingredients necessary for subjects,
% we will then describe frames in
% Section ?? and how they, too, can be disassembled and
% reassembled as needed.
% Finally, Section ?? will give the punchline.
% \note{revisit this paragraph and rework to more explicitly
% note subject/frame organization, also mention zipping}

% In particular, they have a \emph{bottom-up} zipper structure---consisting
% of a focused substructure $\subject$ and, separately, its surrounding
% ``inside-out'' context $\zframe$---

% \note{add some high level preview about edit states, zippers,
% justify this organization so that people are happy to wait;
% consider giving an example right away as I discuss bottom-up zippers}

% Parsing in \ty~ is stratified into two phases:
% assembling shards into tiles, and precedence-parsing
% tiles into terms.
% We will begin in Section \ref{sec:terms-tiles-tips}
% by showing that the precedence-parsing phase is
% guaranteed to succeed provided each element of the
% input sequence ``fits'' its neighbors.
% \note{sort of misleading, ignores children}
% As a result, we can structure \ty~ edit states using
% tiles rather than terms (Section \ref{sec:edit-states}),
% and the actions operating on edit states need only
% concern themselves with shard assembly and fixing holes
% to maintain sequential ``fit'' (Section \ref{sec:actions}).

% \note{something about how both of these parsing steps
% deviate from the usual approach of parsing, which takes
% a sequence of characters or tokens and produces a
% parse tree, whereas here we have more fine-grained
% delineation between different parseable/parsed structures,
% which is why we take care in this section to define
% derivations from one structure to another}

\subsection{Shards and pieces} \label{sec:shards-and-pieces}

Given below is the syntax for shards corresponding to the tiles
in Figure \ref{fig:tile-syntax}.
\input{shard-syntax}

\noindent \note{give an example of shard, note that
we don't model all shards that show up in overview and
that we'll discuss this when we talk about extensions}
Collectively we refer to shards and tiles as \emph{pieces};
a sequence of pieces is called a \emph{segment}.
\input{pieces-syntax}

% \noindent
% \note{something about pieces being quite flexible, forward
% referencing Section \ref{sec:sec:subjects-and-frames} and
% nothing how they make it easy to encode partial selections
% as well as intermediate disassembled edit states
% between input and output edit states of an action}

\input{subject-disassembly}

We can produce shards from tiles via \emph{step-disassembly},
as defined for pieces and segments in Figure \ref{fig:subject-disassembly}.
The first judgment $\pieceDisassembles{\selem}{\selection}$
step-disassembles a piece $\selem$ into a segment of
its constituent shards and children tiles.
The second judgment $\stepDisassembleSelection{\selection_1}{\selection_2}$
chooses a piece in $\selection_1$ and replaces it with its
step-disassemblage to produce $\selection_2$.
The reflexive transitive closure $\tpiecesdd^*$ of
$\tpiecesdd$ we call \emph{disassembly}.
\note{give examples from overview}

Disassembly of segments is analogous to string derivation for a
context-free grammar (CFG),
where a string containing terminal and nonterminal symbols is iteratively
rewritten according to the grammar rules.
A parser for a CFG may then be characterized as a function
that takes a string of terminal symbols and identifies,
if any exist,
a derivation from the grammar's starting nonterminal
symbol to the input string.
We may similarly characterize a parser on pieces according
to the disassembly relation.
In this case, however, our parsing goal is maximal,
not maximum, structure, \ie, the parser
need not identify a disassembly connecting the input
shards to unbroken tiles (the start symbol), but need only assemble shards
together opportunistically.
\note{too hand-wavey, come back and revise}

More precisely, we can show that disassembly is a
partial order, and that every segment in this order
has a unique maximal segment that disassembles to it.
\begin{lemma}
  $\tpiecesdd^*$ is a partial order.
\end{lemma}
\begin{lemma}\label{lemma:unique-parsed-selection}
  For every segment $\selection$, there exists a unique
  segment $\parseSelection{\selection}$ such that:
  \begin{itemize}
  \item $\parseSelection{\selection}\searrow^*\selection$, and
  \item $\parseSelection{\selection}$ is maximal: if $\selection'\searrow^*\parseSelection{\selection}$ then $\selection' = \parseSelection{\selection}$.
  \end{itemize}
\end{lemma}
\noindent
We say that the operator $\parseSelection{\cdot}$ performs \emph{segment assembly}.
We defer presenting a constructive implementation of segment
assembly to the yet-imaginary appendix.

% These results allow us to specify, in an algorithm-independent
% fashion, a well-defined \emph{assembly} function.
% \begin{definition}
%   Let $\parseSelection{\selection}$ be
%   the unique maximal segment that disassembles to segment $\selection$.
% \end{definition}
% \noindent
% We defer presenting a constructive implementation of assembly
% to the imaginary appendix.


\subsection{Terms, tips, and holes} \label{sec:tips-and-terms}

Segments form core data structures in \ty~ zippers,
as we will see in Section \ref{sec:subjects-and-frames}.
Their syntactic flexibility simplifies how we define
zipper actions in Section
\ref{sec:actions}, but it also means we need additional
metasyntactic assurances to know they can be parsed into
well-formed program term.

We characterize assembly of pieces into terms
similarly to how we characterized segment assembly
in Section \ref{sec:shards-and-pieces}.
Figure \ref{fig:flatten-term} defines the \emph{term
disassembly} judgments $\flattenTerm{p}{\selection}$
and $\flattenTerm{e}{\selection}$, which decompose
the terms $p$ and $e$ into segments $\selection$.

\input{term-flattening}

The segments produced by term disassembly always satisfy
a couple properties. First, they are always \emph{intact}:
\begin{definition}
  A segment $\selection$ is called \emph{intact} if it consists
  solely of tiles.
\end{definition}
\begin{lemma}
  If $\flattenTerm{p}{\selection}$ or $\flattenTerm{e}{\selection}$ then $\selection$ is intact.
\end{lemma}
\note{call back to initial \tylr~ figure showing term
vs tile view of lambda term}

Second, each piece in a term disassemblage ``fits''
its neighbors, \ie, each piece's left and right \emph{tips}
match, respectively, the right tip of its left neighbor
and the left tip of its right neighbor.
We have suggestively notated the tile and shard forms with
approximations of their tips, but have elided some relevant
sort information.
More completely, we define in Figure \ref{fig:piece-tips}
the syntax for tips and the operators $\leftTip{c}$
and $\rightTip{c}$ that return the left and right tips
of a piece $c$.
\note{describe tips, discuss a tile example and
a shard example}

In fact, this notion of ``fit'' extends to all disassemblages
of a disassembled term. \note{todo: different names for term
disassembly vs tile disassembly}
More precisely, Figure \ref{fig:tip-tip-connected} defines
\emph{$\tau_1\tau_2$-connectedness} of a segment.
\note{a little more discussion, example of connected vs not connected segment}
We say that a segment $\selection$ is \emph{connected} if there exist
tips $\tau_1$ and $\tau_2$ such that $\selection$ is $\tau_1\tau_2$-connected.

\begin{lemma}
  If $\flattenTerm{p}{C}$ then $C$ is $\lltip{\pat}~\rrtip{\pat}$-connected.
  If $\flattenTerm{e}{C}$ then $C$ is $\lltip{\expr}~\rrtip{\expr}$-connected.
  \note{todo: sort-indexed terms}
\end{lemma}

\input{piece-tips}
\input{tip-tip-connected}

These two properties of segments---intactness
and connectedness---are in fact also \emph{sufficient} to
guarantee there exists a term that disassembles to
the segment in question.

\begin{theorem} \label{thm:term-parseability}
  A segment $\selection$ is intact and $\lltip{s}~\rrtip{s}$-connected
  if and only if there exists a term $x$ of sort $s$ such that
  $\flattenTerm{x}{\selection}$.
  \note{todo: sort-indexed terms}
\end{theorem}
\noindent
We call such a segment an \emph{opseq of sort $s$}.

\paragraph{Hole fixing}
We have seen in Section \ref{sec:overview} how tile-based
editing supports temporary departure from intactness of
the edit state while selecting and restructuring.
However, edit actions must take care to maintain
connectedness at all times to preserve opseq structure
upon reassembling shards and returning to pointing mode.

\input{hole-fixing-judgment}

Toward this end, we will see that actions make use of the hole fixing judgment
$\fixHolesSelections{\selection_1}{\selection_2}{s}{\selection'_1}{\selection'_2}$
given in Figure \ref{fig:fixholes-2},
which takes a pair of segments $\selection_1$ and $\selection_2$
and inserts and removes holes of sort $s$ to produce $\selection'_1$
and $\selection'_2$.
\note{add discussion of auxiliary judgment}
Hole fixing is governed by the following lemma.

\begin{lemma}
  Suppose $\selection_1$ and $\selection_2$ are connected segments
  and $\fixHolesSelections{\selection_1}{\selection_2}{s}{\selection'_1}{\selection'_2}$.
  Then,
  \begin{enumerate}
    \item $\selection'_1\selection'_2$ is $\lltip{s}\rrtip{s}$-connected;
    \item $\selection'_1$ and $\selection'_2$ consist of the same non-hole pieces as
        $\selection_1$ and $\selection_2$ respectively; and
    \item there exist no neighboring holes in $\selection'_1\selection'_2$.
  \end{enumerate}
  \note{formalize 2 and 3}
\end{lemma}

\subsection{Subjects and frames} \label{sec:subjects-and-frames}
Zippers $\editState = \zipper{\subject}{\zframe}$ decompose
expression opseqs into a focused subject $\subject$ and
its surrounding frame $\zframe$.

\paragraph{Subjects}
Prior zipper designs typically model their subjects
as direct instances of the total structure they decompose,
\eg, the subject of a tree-based zipper is a tree.
In our case, this would mean modeling our subject as
an opseq.
However, in order to support text-like editing,
where the cursor sits between characters, we
additionally decompose the subject opseq into
two or more segments as dictated by the cursor position
and editing mode.
% Subjects are \emph{modal},
% each variant corresponding respectively
% to the pointing, selecting, and restructuring modes we
% described informally using \tylr~ in Section \ref{sec:overview}.
\[\arraycolsep=4pt\begin{array}{rlrl}
  \text{subject} & \subject & ::= &
  \pointing{\selection}{\selection} ~\vert~
  \selecting{\selection}{\selection}{\selection} ~\vert~
  \restructuring{\selection}{\selection}{\selection}
\end{array}\]
The three variants above correspond respectively to the pointing,
selecting, and restructuring modes we described
informally using \tylr~ in Section \ref{sec:overview}.
\note{give examples of subjects in each mode from overview}

We do not aim to model the full functionality of \tylr~
described in Section \ref{sec:overview}.
For example, note that we restrict selecting mode so that
a segment can be selected right-to-left but not left-to-right.
We also restrict restructuring mode to carry a single
segment in the backpack at a time, unlike the multi-selection
restructuring we observed in Section ??.
These additional features do not require any new fundamental
reasoning, as we will discuss in Section ??.

% \input{edit-state-syntax}

\paragraph{Frames}
Corresponding to the ``unzipped'' portion of the zipper,
frames model the remainder of the program not included
in the subject, the ``zipped'' portion.
Frames may be understood as a \emph{bottom-up} sequence of
nested containers, where the head container immediately
wraps the subject, and each container points to its
parent container, up to the program root.

The syntax for frames is given below.
At the top level, frames $\zframe$ are sort-indexed \emph{tile frames}
$\tfrelem^s$, which represent tiles with a missing child of
sort $s$.
Other than the program root $\froot$, we annotate tile frame literals
with $\framehole$ symbols in place of the missing children.
Each tile frame $\tfrelem^s$ points to its parent
\emph{sequence frame} $\tframe^{s'}$, which carries the tile frame's
sibling tiles of sort $s'$.
Each sequence frame $\tframe^s$ in turn points to its parent
tile frame $\tfrelem^s$.
\note{give an example from overview}

\[\arraycolsep=3pt\begin{array}{rlrl}
  \text{frame} & \zframe & ::= & \tfrelem^{\pat} ~\vert~ \tfrelem^{\expr} \\\\
  \text{sequence frame} & \tframe^s & ::= & \tframelit{\tiles^s\framehole\tiles^s}{\tfrelem^s} \\
  \text{pattern tile frame} & \tfrelem^{\pat} & ::= &
    \tframelit{\sparen{\framehole}}{\tframe^{\pat}} \\
  & & \vert &
    \tframelit{\tlam{\framehole}}{\tframe^{\expr}} \\
  & & \vert &
    \tframelit{\tlet{\framehole}{\tiles^{\expr}}}{\tframe^{\expr}} \\
  \text{expression tile frame} & \tfrelem^{\expr} & ::= &
    \froot \\
  & & \vert &
    \tframelit{\sparen{\framehole}}{\tframe^{\expr}} \\
  & & \vert &
    \tframelit{\tlet{\tiles^{\pat}}{\framehole}}{\tframe^{\expr}}
    % \scond{}{\tiles^{\expr}}{}
\end{array}\]

% \subsubsection{Tiles, tokens, \& selections}
% The subject $\sigma$ of a zipper $\zipper{\subject}{\tfrelem^s}$
% may be in one of three modes.
% \note{name them here}
% We defer higher-level discussion regarding their roles in
% editing to Section \note{ref}, where we define \ty's editing
% operations, and focus now on the structures that constitute them.

% The first mode, called \emph{pointing mode}, consists of
% a pair of tiles of the sort expected by the frame.
% We described the syntax of tiles
% in Section \ref{sec:overview};
% the same syntax that was presented in Figure \note{ref}
% is presented again in Figure \note{ref} in context.

% The second and third modes, called \emph{selecting mode}
% and \emph{restructuring mode}, each consist of a trio of
% \emph{selections}.

% \note{add back some discussion of tiles/tokens here}

% All three modes consist of two or more \emph{selections}.
% A selection is a linear sequence of heterogeneously
% sorted \emph{tiles} and \emph{tokens}.
% We described the syntax of tiles $\tile^s$
% in Section \ref{sec:overview};
% the same syntax that was presented in Figure \ref{fig:tile-syntax}
% is presented again in Figure \ref{fig:edit-state-syntax} in context.
% Tokens $\shard^s$ are the lexical components of tiles, forming either
% the substance of childless tiles or the delimiters of parent
% tiles' children.
% Tokens are generated from tiles via the \emph{tile disassembly} function
% $\disassembleTile{\cdot}$, defined in Figure \ref{fig:disassemble-tile},
% which takes a tile and produces
% a selection consisting of the tile's tokens
% and the tile's children tiles.
% \note{need more examples}

% \note{find different phrasing here that doesn't invite interpretation as technical terms}
% Selections have a permissive syntax, with multiple
% ways of representing the same essential content.
% For example, both a singleton tile and the tile's disassembly
% form valid selections.
% \note{make this a real example}
% Such a permissive syntax is useful for representing
% arbitrary divisions of a program's
% lexical components, but we would like to ensure
% that the divided selections remain ``maximally assembled''
% in order to inform the user of the top-level structures
% they are manipulating.
% \note{give explicit examples of divisions showing maximally
% and non-maximally assembled selections}

% This calls for defining an appropriate
% parser on selections to maintain
% maximal structure as they grow and shrink during editing.
% \note{make this more precise "when the selection changes",
% perhaps add another example here of a selection growing}
% Such a parser cannot be specified in the standard way
% with a context-free grammar because such parsers are
% all-or-nothing, whereas in our case our parser should
% produce structure opportunistically while not requiring any
% specific top-level form.
% \note{maybe don't talk about this here, if anything talk about
% it after I define what we did}
% Much as a CFG-based parser identifies a derivation from
% the CFG's start symbol to the input string,
% our selection parser identifies an analogous ``disassembly'' from
% a maximally structured selection to the input selection.
% \note{more example}

\subsubsection{Frames \& zippers}
The sort-indexed frame $\tfrelem^s$ of a zipper
$\zipper{\subject}{\tfrelem^s}$ models the rest of
the program surrounding the subject $\subject$,
the sort index $s$ specifying what sort is expected
of the subject.
Frames are nested in bottom-up fashion, starting
with the nearest containing structure and concluding
with the program root.
For example, if the edit subject is located at $\Box$
in the program \texttt{let x = ( $\Box$ ) + 1 in x + 2},
then the frame is represeted in our syntax as
\[
  \tframelit{
    \sparen{\_}
  }{
    \tframelit{
      \_\splus{}{\texttt{1}}
    }{
      \tframelit{
        \slet{\texttt{x}}{\_}{}
      }{
        \tframelit{
          \_\ \splus{\texttt{x}}{\texttt{2}}
        }{
          \note{root}
        }
      }
    }
  }
\]
Notice that frames alternate between \emph{tile
frames} $\tfrelem^s$ that form bidelimited containers and
\emph{sequence frames} $\tframe^s$ consisting of the tiles
surrounding each tile frame.

% Having defined subjects and frames, we may now
% define the syntax of edit states,
% shown in Figure \note{ref}.
% An edit state consists of a sort-indexed \emph{zipper},
% each consisting of a subject and a frame of the
% specified sort.

% \input{frame-syntax}
% \input{zipper-syntax}
\input{disassemble-frame-tile}
\input{step-disassemble-zipper}

Like tiles, frames may be decomposed into their
lexically constituent tiles and tokens via the
\emph{frame disassembly} function $\disassembleTileFrame{\cdot}$,
defined in Figure \note{ref}.
Similar to how we lift tile disassembly to selection
disassembly, we can lift frame disassembly to
\emph{edit state step-disassembly}, defined in Figure \note{ref}.
We can show that the reflexive transitive closure
$\nearrow^*$ of $\nearrow$, which we call \emph{edit state disassembly},
is a partial order in
which every edit state with a subject in pointing mode
has a unique minimal element that disassembles to it:
\begin{lemma}
  $\nearrow^*$ is a partial order.
\end{lemma}
\begin{lemma}\label{lemma:unique-parsed-editstate}
  For every zipper $\editState = \zipper{\pointing{\selection_1}{\selection_2}}{\tfrelem^s}$
  in pointing mode,
  there exists a unique zipper $\parseZipper{\editState}$ such that
  \begin{itemize}
  \item $\parseZipper{\editState} \nearrow^* \editState$, and
  \item $\parseZipper{\editState}$ is minimal: if $\editState' \nearrow^* \parseZipper{\editState}$ then $\editState' = \parseZipper{\editState}$.
  \end{itemize}
\end{lemma}
\noindent
We say that the operator $\parseZipper{\cdot}$ performs \emph{zipper assembly}.

% \begin{definition}
%   $\zeta^s_1\nearrow\zeta^s_2$ if $\disassembleTileFrame{\zeta^s_1} = \zeta^s_2$.
%   \note{defined this for symmetry with diassembling of tiles}
% \end{definition}

% \begin{definition}
%   $\nearrow^*$ is the reflexive transitive closure of $\nearrow$.
% \end{definition}

\subsection{Actions} \label{sec:actions}

We now present \ty's central action judgment
$\performAction{\editState_1}{\action}{\editState_2}$,
which sends edit state $\editState_1$ to $\editState_2$
via action $\action$.
The action judgment is governed by a syntactic sensibility
theorem (Theorem \note{ref}) that ensures that
the subject of every action-reachable edit state can
be parsed into language term.

\input{filter-tiles}
\input{whole-selection}
\input{action-syntax}

\input{move-pointing}
\input{move-selecting}
\input{move-restructuring}
\input{action-judgment}

\note{
  Define actions more generally than \tylr~ implementation,
  such that delimiters are assigned sorts to their ends,
  selections are not restricted to shards and
  same-sort tiles, and only selections whose ends are the
  same may enter restructuring mode.
}

% (2 * [3])  ->  (2 * (3 + [_]))

\begin{itemize}
  \item explain action judgment + choice rules
  \item theorems
  \begin{itemize}
    \item movability
    \item selectability
    \item restructuring is sound and complete
  \end{itemize}
\end{itemize}

\begin{theorem}
  Suppose $Z = \zipper{\subject}{\tfrelem^s}$ is a reachable edit state.
  \begin{enumerate}
  \item[(1)] If $\subject = \pointing{\selection_1}{\selection_2}$,
    then $\selection_1\selection_2$ is an opseq of sort $s$.
  \item[(2)] If $\subject = \selecting{\selection_1}{\selection_2}{\selection_3}$,
    then $\parseSelection{\selection_1\selection_2\selection_3}$ is an opseq of sort $s$.
  \item[(3)] If $\subject = \restructuring{\selection_1}{\selection_2}{\selection_3}$...
    \begin{enumerate}
      \item[(a)] ...and $\selection_2$ is intact, then
        $\selection_1\selection_3$ is an opseq of sort $s$.
      \item[(b)] ...and $\selection_2$ is cracked, then
        $\parseSelection{\selection_1\selection_2\selection_3}$ is an opseq of sort $s$.
    \end{enumerate}
  \end{enumerate}
\end{theorem}

\begin{theorem}

\end{theorem}

% \subsection{Typechecking}
% \note{
%   Define tree-structured variants of tiles and frames.
%   Define precedence parser that converts linear forms to tree forms
%   (or assume existence of such parser?).
%   Define type
% }
